Astrophysical calibration of gravitational-wave detectors

### 2015-08-07 docs/cosmic_cal.tex
Some more updates to the document

\newcommand{\curlH}{\mathcal{H}}
\newcommand{\gws}{\tilde{h}}

@@

% acronym definitions
\acrodef{GW}[GW]{gravitational wave}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivalent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{NSBH}[NSBH]{neutron star-black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}
\acrodef{sGRB}[sGRB]{short gamma-ray burst}
\acrodef{CBC}[CBC]{compact binary coalescence}
\acrodef{EM}[EM]{electro-magnetic}
\acrodef{aLIGO}[aLIGO]{Advanced LIGO}
\acrodef{AdV}[AdV]{Advanced Virgo}
\acrodef{PSD}[PSD]{power spectral density}

@@

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$, changes of the 
interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is the full arm length. Calibration is 
required to relate the actual measured interferometer error signal output $e(f)$ to $\Delta L$. 
This relation is known as the length response function, $R(f)$, defined such that
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
where where we assume $R$ varies only slowly in time (in comparison to transient signal 
time-scales). Calculation of $R$ requires the measurement of various functions within a control 
feedback loop (see \cite{2010NIMPA.624..223A}, which are subject to uncertainties. In this 
study we will assume an estimate of $R$ is available (although in theory we could take on the role 
of estimating $R$ itself), but that it differs from the truth through some unknown scale factor, 
$s$, so that
\begin{equation}\label{eq:scalefactor}
sh(f) = h_{\rm m}(f) = \frac{R(f,t) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $s > 1$ then a signal would appear to have a larger amplitude (e.g.\ be closer) than 
in reality\footnote{However, the \ac{SNR} would be the correct value as the signal and the noise 
will both contain the scale factor}, whereas if $s < 1$ it would appear to have a lower amplitude 
than reality. In this analysis we will simplify the situation by assuming that $s$ is a constant 
with respect to frequency, but in future studies that assumption could be dropped and $s$ could 
take some functional form, or piece-wise fit, with respect to $f$ (e.g.\ in a similar way to the 
method of \cite{2013PhRvD..88h4044L} used for fitting differences in \ac{PSD} estimates).

@@

% Define the likelihood function we will be using
We consider that the measured data in any detector is defined as
%
\begin{equation}
  \tilde{d}_{k}(f)=s(\tilde{n}_{k}(f)+\gws_{k}(f,\btheta))
\end{equation}
%
where $\tilde{n}(f)$ is the Fourier transform of the true (Gaussian) strain noise drawn from a 
distribution consistent with the \ac{PSD} of the advanced detector noise and $\btheta$ is the set 
of waveform parameters.

@@

To assess the ability to estimate the unknown calibration scale factors we want to calculate their 
marginal posterior probability distribution functions (pdf). We use Bayes' theorem for which we 
need to define a likelihood function and prior pdfs on the signal parameters being estimated. In 
this analysis we use the Markov chain Monte Carlo code {\tt emcee} \cite{2013PASP..125..306F} to 
sample the posterior distribution. We use a Gaussian likelihood function given by
%
\begin{equation}
\fl p(\bi{d} | \btheta, \bi{s}, \curlH, I) \propto \exp{\left[4\Delta f \sum_{k=1}^{N_{\rm det}}
  \sum_{i=i_{f_{\rm low}}}^{i_{f_{\rm high}}}\frac{\Re{\left\{ \tilde{d}^{*}_{k,i} 
\gws_{k,i}(\btheta) - \frac{1}{2}\left(\gws^*_{k,i}(\btheta)\gws_{k,i}(\btheta) + 
\tilde{d}^*_{k,i}\tilde{d}_{k,i}\right) \right\}}}{S_{k,i}}\right]}
\end{equation}
% 
where $\bi{d}$ is an array containing the data for all detectors, $\Delta f$ is the frequency 
resolution, $S_k$ is the $k^{\rm th}$ detector's one sided noise \ac{PSD}, and the $i$ indices 
increment over frequency with $i_{f_{\rm low}}$ and $i_{f_{\rm high}}$ corresponding to the lower 
and upper range in frequencies used for our analysis, which were 20\,Hz to 400\,Hz respectively. 
This frequency range was chosen as the vast majority of the \ac{SNR} for the inspiralling signals 
we use can be found within this range, so going to lower or higher frequencies provides very little 
additional information. The source parameters, and the unknown scale factors, have been subsumed 
into a vector $\btheta$.

@@

The primary reason why we can use coincident observations with a \ac{sGRB} as a potential check of 
detector calibration is that the \ac{GRB} allows us to constrain the priors on various parameters 
that would normally be highly correlated with the signal amplitude. The most important of these is 
that the \ac{GRB} can provide a very tight constraint on the source distance\footnote{For \ac{BNS} 
and \ac{NSBH} systems they will be observable only in the relatively nearby Universe ($z \sim 
0.1$ and $z\sim 0.2$ for \ac{BNS} and \ac{NSBH} systems respectively), so the relation between the 
measured GRB redshift and distance is simple and our luminosity distance will be negligibly 
effected by the choice of cosmology \MP{Check this?}.}. In this analysis we therefore assume that 
the source distance is known (i.e.\ has a $\delta$-function prior). Another piece of information 
that we can use from having a coincident \ac{GRB} is that the system is likely to be relatively 
close to face-on (an therefore nearly circularly polarised), which allows us to place prior 
constraints on the system inclination angle. Beam opening angles for \acp{sGRB} are relatively 
poorly constrained as they are based on only a few jet break detections, however 
\cite{2014ApJ...780..118F} give a median opening angle value of $\sim 10^{\circ}$. If we assume 
that the beam opening angle is a half-normal distribution, for which the median is given by 
$\sqrt{2}\erf{}^{-1}(1/2)\sigma \sim 0.67\sigma$, then a median of $\sim 10^{\circ}$ corresponds to 
$\sigma \sim 14.8^{\circ}$. We therefore place a Gaussian prior on the inclination angle with zero 
mean and $\sigma=14.8^{\circ}$.

@@

The final important piece of information that we can make use of from the GRB observation is the 
sky position of the source, which we take to be known precisely.

For the time of coalescence we use a uniform prior of spanning $\pm0.01$\,seconds around the 
recorded time of the signal that would be returned by the detection pipeline. For the reference 
phase a uniform prior between 0 and $2\pi$ is used, and for the polarisation angle a uniform prior 
between 0 and $\pi/2$ is used. As the signals we use are generally close to being circularly 
polarised (face-on) the phase and polarisation angle will be degenerate meaning the exact 
combination of them will have very little effect of the result.

Finally, we require a prior on the calibration scale factors for each detector. We make the 
assumption that in general the calibration applied to the detector data will be correct, so want to 
use a prior that is peaked at unity. We also assume that the probability density that the 
calibration scale factor is either larger or smaller than unity by an equivalent factor, e.g.\ 
$s=10$ or $s=0.1$, is the same. For the scale factor for each detector we we therefore use a 
log-normal distribution as our prior
\begin{equation}
 p(s|I) = \frac{1}{s\sigma\sqrt{2\pi}}\exp{\left( -\frac{(\ln{s} - \mu)^2}{2\sigma^2} \right)},
\end{equation}
where we chose $\sigma = 1.07$, which in turn means that having the mode at unity gives $\mu = 
1.15$. With these parameters the prior probability density at 0.1 and 10 is one tenth of that at 
1. Note that this prior does not give a symmetric amount of probability about unity, but given our 
simulation criterion described below in Section~\ref{sec:simulations} we find that the likelihood 
generally overwhelms the prior. As $s$ is a scale factor we also have the further constraint that 
it must be positive.

@@

% Describe the ensemble of simulations used
Separately for the \ac{BNS} and \ac{NSBH} systems we have performed a large number simulations at 
each distance value. For proposing injections for each simulation the source parameters were 
randomly drawn from the prior distributions given in Section~\ref{sec:priors}, with the sky 
position being drawn randomly from a uniform distribution, the coalescence times held fixed within 
the centre of the prior window, and with the calibration scaling factors for each detector drawn 
from a Gaussian distribution with a mean of one and standard deviation of $0.125$ (equivalent to a 
mean fractional offset of 10\%). However, in accepting a proposed injection as one to be analysed 
we introduced a criteria that the signal be ``detectable'', based on that used in 
\cite{2012PhRvD..85h2002A} of it having an \ac{SNR} of $\geq 5.5$ in at least two detectors (as 
these would be signals coincident with a \ac{sGRB} we do not include the often used further 
constraint that the total coherent network \ac{SNR} is greater than 12). This criteria means that 
at larger distances our population of injected sources was not uniformly distributed over the sky, 
and were generally closer to being circularly polarised.

These simulations have allowed us to assess how well on average we would be able to constrain the 
calibration scale for a given \ac{CBC}-\ac{GRB} coincidence.

@@

For each of the simulated sources we have used the {\tt emcee} python MCMC package 
\cite{2013PASP..125..306F} to perform parameter estimation over the unknown source parameters using 
the priors as discussed in Section~\ref{sec:priors}. In each case when calculating the likelihood 
we used an estimate of the noise \ac{PSD} based on the advanced detector design sensitivities 
(using those given in \cite{2013arXiv1304.0670L}), but simulated as if calculated by averaging 64 
separate noisy \ac{PSD} estimates and scaled with the same factor as applied to the injection and 
noise\footnote{We do not account for there being a potential difference between the estimated 
\ac{PSD} and the actual \ac{PSD} of the analysed section of data as described in e.g.\ 
\cite{2013PhRvD..88h4044L}. This difference would be very highly correlated with the calibration 
scale factor, so in reality our estimate of the scale factor would be a combination of the
calibration offset and the difference in the \ac{PSD}. As such our results on real data would be an 
upper limit on the calibration scale factor.}. This has provided posterior probability distributions 
on the calibration scale factors for each detector. Examples of the marginalised posterior 
probabilities for a \ac{BNS} system and a \ac{NSBH} system observed with the three detector network 
are shown in figures~\ref{fig:bnspost} and \ref{fig:nsbhpost} respectively.

From these posterior distributions we have calculated the 68\% credible region on the calibration 
scale factors for each detector (if these were Gaussian distributions this is equivalent to the 
region either side of the mean bounded by the $1\sigma$ intervals). For all the signals at each 
distance increment we have produced the distribution of the fractional half widths (i.e.\ $1\sigma$) 
of these scale factors' confidence intervals compared to the actual value. These are shown as 
boxplots in figures~\ref{fig:bnsresults} and \ref{fig:nsbhresults} for the \ac{BNS} and \ac{NSBH} 
systems respectively. The boxes show the extent from the lower to upper quartile of the values, 
whilst the whiskers extend from the 5$^{\rm th}$ to 95$^{\rm th}$ percentile. The black line 
within each box gives the median value and the star gives the mean value. Also shown on each plot as 
the dashed magenta line is the percentage of signals drawn from the prior distribution that fulfill 
the detection criterion.

@@

In figure~\ref{fig:bnsresults} we see that on average the scale factor can be recovered to 
equivalent precision in both \ac{aLIGO} detectors, as would be expected, with uncertainties 
generally within the 10\% range for sources at 100\,Mpc. This is comparable to previous estimates 
of the calibration error in the initial LIGO detectors. An interesting feature is that for 
distances $\gtrsim 250$\,Mpc the upper extent of the uncertainty for H1 and L1 hits a maximum at 
$\sim 25\%$, whilst the width of boxes narrows. This is due to our ``detectability'' criteria, 
in that for all distances we will only be seeing those sources with a high enough \ac{SNR} that we 
would consider them detectable. There will therefore be fewer sources with \ac{SNR} higher than 
this criteria at large distances, giving us a narrower range, and we automatically exclude those 
with \ac{SNR} that are too small thus truncating our uncertainty distribution at the upper end. 
However, this does show that on average for sources that are detectable out to 450\,Mpc we would be 
able to constrain the calibration scale factor uncertainty for the \ac{aLIGO} detectors to $\sim 
20\%$. As the detection criteria will generally come from the \ac{SNR} being high enough within the 
two \ac{aLIGO} instruments this means that the SNR in \ac{AdV} can be small and thus the ability to 
constrain the scale factor for it becomes poor (although it still provides information that the 
calibration is not grossly inaccurate). We also see that true uncertainties achievable for 
recovering the calibration scale factors are indeed very similar to the simple assumption that they 
would be given by $sim 1/{\rm SNR}$, i.e.\ at the upper end of the distribution the \ac{SNR} in 
the two \ac{aLIGO} detectors will be $\sim 5.5$, which would be expected to produce uncertainties 
of $\lesssim 20\%$. Our results are probably slightly poorer than the \ac{SNR} expectation due to 
their still be correlations with $\iota$ even within the constrained prior range, and slight 
correlations between the scale factors for H1 and L1 as seen in figure~\ref{fig:bnspost}.

In figure~\ref{fig:nsbhresults} we see very similar results for the \ac{NSBH} systems although the 
higher \ac{SNR} of the signals means that we can provide $\sim 20\mbox{--}25\%$ uncertainties on 
the calibration scale factors for H1 and L1 out to greater distances. The distributions generally 
appear slightly broader than those for the \ac{BNS} systems. This may be entirely explicable 
through two factors. The first factor down to the fact that including the spin parameter leads to 
string correlations between the chirp mass, mass ratio and spin. These strong correlations make 
convergence of the MCMC take longer and means we have fewer independent samples to estimate the 
posterior distributions with. This leads to a larger statistical fluctiation on the results. The 
second factor is that there appears to be a modest effect due to the population of sources that 
are detectable. For distances at which the \ac{BNS} and \ac{NSBH} systems would give the same 
percentage of obsverable sources there is a slight increase in the mean and standard distribution of 
\acp{SNR} of the \ac{NSBH} systems over the \ac{BNS} systems.

As noted previously the mass distribution for the black holes could be quite different from the one 
we used, with masses extending to $\gtrsim 10$\,M$_{\odot}$. We expect these would produce similar 
results to those we see in figure~\ref{fig:nsbhresults}, but again extending to further distances. 
Introducing the merger and ring-down phase, which we currently ignore, may also help increase 
\ac{SNR}.

@@

We have attempted to see how well we can assess the detector calibration for \ac{aLIGO} and 
\ac{AdV} using astrophysical sources. To do this we require that a \ac{sGRB}, for which it has been 
possible to measure the distance, is observed in coincidence with a \ac{CBC} signal in the \ac{GW} 
detectors. This enables us to assume a known distance to the source, and also limit the inclination 
of the source, which in turn allows us to test the consistency of the detector calibration. We do 
this by including an unknown scale factor on the true signal detector data, which we estimate given 
the data. We find that for detectable \ac{BNS} sources the uncertainty of on the calibration scale 
factor could on average be determined to $\lesssim 10\%$ of its true value for the \ac{aLIGO} 
and \ac{AdV} detectors out to 100\,Mpc. This is comparible to the hoped for level of calibration 
accuracy of the detectors. For sources at the standard single detector \ac{BNS} horizon distance of 
$\sim 450$\,Mpc the scale factor could on average be determined to within $\lesssim 20\%$ of its 
true value for the \ac{aLIGO} detectors. Similar results were found for \ac{NSBH} sources, although 
sources could be observed out to higher distances.

The requirement of a coincident \ac{sGRB} with a known distance means that there will be 
considerable latency for this calibration assessment, but importantly this method would provide 
semi-independent consistency check that the calibration is reasonably accurate. We do note that this 
method has to make its own assumptions, including that the waveforms we use accurately reflect the 
signal and the calibration, the phase calibration does not change to much over frequency and that 
the overall calibration does not change considerably over the length of a signal. These are studies 
that could be incorporated in the future.

We imagine that a test for consistency of calibration could be performed with an ensemble of 
\ac{CBC} sources, although such a method may have to marginalise over highly uncertain population 
models and may give quite weak bounds on the overall calibration.

We emphasise that our work has assumed that the PSD used in the likelihood function
when estimating the calibration scale factor is an accurate representation of the
PSD at the time of the observed CBC signal. In reality the PSD may be estimated from
a period close to, but not overlapping, the signal time and therefore may be slightly
different \cite{2013PhRvD..88h4044L}. Our result would therefore be completely correlated
with any uncertainty in the PSD estimate, but would still offer an upper limit
on the calibration uncertainty.

A further extension of this work would be for the calibration scale factor to be estimated as a 
function of frequency. This would require the frequency series to be broken up into a set of 
sections and the calibration scale factor estimated for each. However, the scale factor uncertainty 
will obviously be related to the \ac{SNR} of a signal within a particular section. It could 
therefore be a job of Bayesian model selection to work out the optimal number of frequency 
sections to estimate the scale factor for. The BayesLine algorithm \cite{2015PhRvD..91h4034L}, or 
the method described in \cite{2013PhRvD..88h4044L}, may provide a natural way to perform such an
analysis. Indeed these methods, and those of \cite{Vitale:2012} are already being used for 
marginalising over calibration uncertainties, but do not attempt to estimate it.

>>>>

\newcommand{\curlH}{\mathcal{H}}
\newcommand{\gws}{\tilde{h}}
\newcommand{\scf}{\ensuremath{\mathcal{C}}}

@@

% acronym definitions
\acrodef{GW}[GW]{gravitational wave}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivalent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{NSBH}[NSBH]{neutron star-black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}
\acrodef{sGRB}[sGRB]{short gamma-ray burst}
\acrodef{CBC}[CBC]{compact binary coalescence}
\acrodef{EM}[EM]{electro-magnetic}
\acrodef{aLIGO}[aLIGO]{Advanced LIGO}
\acrodef{AdV}[AdV]{Advanced Virgo}
\acrodef{PSD}[PSD]{power spectral density}
\acrodef{MCMC}[MCMC]{Markov chain Monte Carlo}

@@

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$, changes of the 
interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is the full arm length. Calibration is 
required to relate the actual measured interferometer error signal output $e(f)$ to $\Delta L$. 
This relation is known as the length response function, $R(f)$, defined such that
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
where where we assume $R$ varies only slowly in time (in comparison to transient signal 
time-scales). Calculation of $R$ requires the measurement of various functions within a control 
feedback loop (see \cite{2010NIMPA.624..223A}, which are subject to uncertainties. In this 
study we will assume an estimate of $R$ is available (although in theory we could take on the role 
of estimating $R$ itself), but that it differs from the truth through some unknown scale factor, 
$\scf$, so that
\begin{equation}\label{eq:scalefactor}
\scf h(f) = h_{\rm m}(f) = \frac{R(f,t) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $\scf > 1$ then a signal would appear to have a larger amplitude (e.g.\ be 
closer) than in reality\footnote{However, the \ac{SNR} would be the correct value as the signal and 
the noise will both contain the scale factor}, whereas if $\scf < 1$ it would appear to have a 
lower amplitude than reality. In this analysis we will simplify the situation by assuming that 
$\scf$ is a constant with respect to frequency, but in future studies that assumption could be 
dropped and $\scf$ could take some functional form, or piece-wise fit, with respect to $f$ (e.g.\ 
in a similar way to the method of \cite{2013PhRvD..88h4044L} used for fitting differences in 
\ac{PSD} estimates).

@@

% Define the likelihood function we will be using
We consider that the measured data in any detector is defined as
%
\begin{equation}
  \tilde{d}_{k}(f)=\scf_{k}(\tilde{n}_{k}(f)+\gws_{k}(f,\btheta))
\end{equation}
%
where $\tilde{n}(f)$ is the Fourier transform of the true (in our case Gaussian) strain noise drawn 
from a, $\btheta$ is the set of waveform parameters and $\scf_k$ is the calibration scale factor for 
the $k^{\rm th}$ detector.

@@

To assess the ability to estimate the unknown calibration scale factors we want to calculate their 
marginal posterior probability distribution functions (pdf). We use Bayes' theorem for which we 
need to define a likelihood function and prior pdfs on the signal parameters being estimated. In 
this analysis we use the Markov chain Monte Carlo code {\tt emcee} \cite{2013PASP..125..306F} to 
sample the posterior distribution. We use a Gaussian likelihood function given by
%
\begin{equation}
\fl p(\bi{d} | \btheta, \vec{\scf}, \curlH, I) \propto \exp{\left[4\Delta f \sum_{k=1}^{N_{\rm 
det}}
  \sum_{i=i_{f_{\rm low}}}^{i_{f_{\rm high}}}\frac{\Re{\left\{ \tilde{d}^{*}_{k,i} 
\scf_{k}\gws_{k,i}(\btheta) - \frac{1}{2}\left(\scf_k^2 \gws^*_{k,i}(\btheta)\gws_{k,i}(\btheta) + 
\tilde{d}^*_{k,i}\tilde{d}_{k,i}\right) \right\}}}{S_{k,i}}\right]}
\end{equation}
% 
where $\bi{d}$ is an array containing the data for all detectors, $\Delta f$ is the frequency 
resolution, $S_k$ is the $k^{\rm th}$ detector's measured one sided noise \ac{PSD} (and as such 
will contain the effect of the calibration scale factor), and the $i$ indices increment over 
frequency with $i_{f_{\rm low}}$ and $i_{f_{\rm high}}$ corresponding to the lower 
and upper range in frequencies used for our analysis, which were 20\,Hz to 400\,Hz respectively. 
This frequency range was chosen as the vast majority of the \ac{SNR} for the inspiralling signals 
we use can be found within this range, so going to lower or higher frequencies provides very little 
additional information. The source parameters have been subsumed into a vector $\btheta$ and the 
calibration scale factors are within the vector $\vec{\scf}$.

@@

The primary reason why we can use coincident observations with a \ac{sGRB} as a check of detector 
calibration is that the \ac{sGRB} allows us to constrain the priors on various parameters that 
would normally be highly correlated with the signal amplitude. The most important of these is 
that the \ac{sGRB} can provide a very tight constraint on the source distance\footnote{For \ac{BNS} 
and \ac{NSBH} systems they will be observable only in the relatively nearby Universe ($z \sim 
0.1$ and $z\sim 0.2$ for \ac{BNS} and \ac{NSBH} systems respectively), so the relation between the 
measured \ac{sGRB} redshift and distance is simple and our luminosity distance will be negligibly 
effected by the choice of cosmology \MP{Check this?}.}. In this analysis we therefore assume that 
the source distance is known (i.e.\ has a $\delta$-function prior). Another piece of information 
that we can use from having a coincident \ac{sGRB} is that the system is likely to be relatively 
close to face-on (and therefore nearly circularly polarised), which allows us to place prior 
constraints on the system inclination angle. Beam opening angles for \acp{sGRB} are relatively 
poorly constrained as they are based on only a few jet break detections, however 
\cite{2014ApJ...780..118F} give a median opening angle value of $\sim 10^{\circ}$. If we assume 
that the beam opening angle is a half-normal distribution, for which the median is given by 
$\sqrt{2}\erf{}^{-1}(1/2)\sigma \sim 0.67\sigma$, then a median of $\sim 10^{\circ}$ corresponds to 
$\sigma \sim 14.8^{\circ}$. We therefore place a Gaussian prior on the inclination angle with zero 
mean and $\sigma=14.8^{\circ}$.

@@

The final important piece of information that we can make use of from the \ac{sGRB} observation is 
the sky position of the source, which we take to be known precisely.

For the time of coalescence we use a uniform prior of spanning $\pm0.01$\,seconds around the 
recorded time of the signal that would be returned by the detection pipeline. For the reference 
phase a uniform prior between 0 and $\pi$ is used, and for the polarisation angle a uniform prior 
between 0 and $\pi/2$ is used. As the signals we use are generally close to being circularly 
polarised (face-on) the phase and polarisation angle will be degenerate meaning the exact 
combination of them will have very little effect of the result.

Finally, we require a prior on the calibration scale factors for each detector. We make the 
assumption that in general the calibration applied to the detector data will be correct, so want to 
use a prior that is peaked at unity. We also assume that the probability density that the 
calibration scale factor is either larger or smaller than unity by an equivalent factor, e.g.\ 
$s=10$ or $s=0.1$, is the same. For the scale factor for each detector we we therefore use a 
log-normal distribution as our prior
\begin{equation}
 p(\scf|I) = \frac{1}{\scf\sigma\sqrt{2\pi}}\exp{\left( -\frac{(\ln{\scf} - \mu)^2}{2\sigma^2} 
\right)},
\end{equation}
where we chose $\sigma = 1.07$, which in turn means that having the mode at unity gives $\mu = 
1.15$. With these parameters the prior probability density at 0.1 and 10 is one tenth of that at 
1. Note that this prior does not give a symmetric amount of probability about unity, but given our 
simulation criterion described below in Section~\ref{sec:simulations} we find that the likelihood 
generally overwhelms the prior. As $\scf$ is a scale factor we also have the further constraint 
that it must be positive.

@@

% Describe the ensemble of simulations used
Separately for the \ac{BNS} and \ac{NSBH} systems we have performed a large number simulations at 
each distance value. When proposing injections for each simulation the source parameters were 
randomly drawn from the prior distributions given in Section~\ref{sec:priors}, with the sky 
position being drawn randomly from a uniform distribution on the sky, the coalescence times held 
fixed within the centre of the prior window, and with the calibration scaling factors for each 
detector drawn from a Gaussian distribution with a mean of one and standard deviation of $0.125$ 
(equivalent to a mean fractional offset of 10\%). However, in accepting a proposed injection as one 
to be analysed 
we introduced a criteria that the signal be ``detectable'', based on that used in 
\cite{2012PhRvD..85h2002A} of it having an \ac{SNR} of $\geq 5.5$ in at least two detectors (as 
these would be signals coincident with a \ac{sGRB} we do not include the often used further 
constraint that the total coherent network \ac{SNR} is greater than 12). This criteria means that 
at larger distances our population of injected sources was not uniformly distributed over the sky, 
and were generally closer to being circularly polarised.

These simulations have allowed us to assess how well on average we would be able to constrain the 
calibration scale for a given \ac{CBC}-\ac{sGRB} coincidence.

@@

For each of the simulated sources we have used the {\tt emcee} python \ac{MCMC} package 
\cite{2013PASP..125..306F} to perform parameter estimation over the unknown source parameters using 
the priors as discussed in Section~\ref{sec:priors}. In each case when calculating the likelihood 
we used an estimate of the noise \ac{PSD} based on the advanced detector design sensitivities 
(using those given in \cite{2013arXiv1304.0670L}), but simulated as if calculated by averaging 64 
separate noisy \ac{PSD} estimates and scaled with the same factor as applied to the injection and 
noise\footnote{We do not account for there being a potential difference between the estimated 
\ac{PSD} and the actual \ac{PSD} of the analysed section of data as described in e.g.\ 
\cite{2013PhRvD..88h4044L}. This difference would be very highly correlated with the calibration 
scale factor, so in reality our estimate of the scale factor would be a combination of the
calibration offset and the difference in the \ac{PSD}. As such our results on real data would be an 
upper limit on the calibration scale factor.}. This has provided posterior probability distributions 
on the calibration scale factors for each detector. Examples of the marginalised posterior 
probabilities for a \ac{BNS} system and a \ac{NSBH} system observed with the three detector network 
are shown in figures~\ref{fig:bnspost} and \ref{fig:nsbhpost} respectively.

From these posterior distributions we have calculated the minimal 68\% credible region on the 
calibration scale factors for each detector (if these were Gaussian distributions this is 
equivalent to the region either side of the mean bounded by the $1\sigma$ intervals). For all the 
signals at each distance increment we have produced the distribution of the fractional half widths 
(i.e.\ $1\sigma$) of these scale factors' confidence intervals compared to the actual value. These 
are shown as boxplots in figures~\ref{fig:bnsresults} and \ref{fig:nsbhresults} for the \ac{BNS} 
and \ac{NSBH} systems respectively. The boxes show the extent from the lower to upper quartile of 
the values, whilst the whiskers extend from the 5$^{\rm th}$ to 95$^{\rm th}$ percentile. The black 
line within each box gives the median value and the star gives the mean value. Also shown on each 
plot as the dashed magenta line is the percentage of signals drawn from the prior distribution that 
fulfil the detection criterion.

@@

In figure~\ref{fig:bnsresults} we see that on average the scale factor can be recovered to 
equivalent precision in both \ac{aLIGO} detectors, as would be expected, with uncertainties 
generally within the 10\% range for sources at 100\,Mpc. This is comparable to previous estimates 
of the calibration error in the initial LIGO detectors. An interesting feature is that for 
distances $\gtrsim 250$\,Mpc the upper extent of the uncertainty for H1 and L1 hits a maximum at 
$\sim 25\%$, whilst the width of boxes narrows. This is due to our ``detectability'' criteria, 
in that for all distances we will only be seeing those sources with a high enough \ac{SNR} that we 
would consider them detectable. There will therefore be fewer sources with \ac{SNR} higher than 
this criteria at large distances, giving us a narrower range, and we automatically exclude those 
with \ac{SNR} that are too small thus truncating our uncertainty distribution at the upper end. 
However, this does show that on average for sources that are detectable out to 450\,Mpc we would be 
able to constrain the calibration scale factor uncertainty for the \ac{aLIGO} detectors to $\sim 
20\%$. As the detection criteria will generally come from the \ac{SNR} being high enough within the 
two \ac{aLIGO} instruments this means that the SNR in \ac{AdV} can be small and thus the ability to 
constrain the scale factor for it becomes poor (although it still provides information that the 
calibration is not grossly inaccurate). We also see that true uncertainties achievable for 
recovering the calibration scale factors are indeed very similar to the simple assumption that they 
would be given by $\sim 1/{\rm SNR}$, i.e.\ at the upper end of the distribution the \ac{SNR} in 
the two \ac{aLIGO} detectors will be $\sim 5.5$, which would be expected to produce uncertainties 
of $\lesssim 20\%$. Our results are probably slightly poorer than the \ac{SNR} expectation due to 
their still be correlations with $\iota$ even within the constrained prior range, and slight 
correlations between the scale factors for H1 and L1 as seen in figure~\ref{fig:bnspost}.

In figure~\ref{fig:nsbhresults} we see very similar results for the \ac{NSBH} systems although the 
higher \ac{SNR} of the signals means that we can provide $\sim 20\mbox{--}25\%$ uncertainties on 
the calibration scale factors for H1 and L1 out to greater distances. The distributions generally 
appear slightly broader than those for the \ac{BNS} systems. This may be entirely explicable 
through two factors. The first factor is from the fact that including the spin parameter leads to 
strong correlations between the chirp mass, mass ratio and spin. These strong correlations make 
convergence of the \ac{MCMC} take longer and means we have fewer independent samples to estimate 
the posterior distributions with. This leads to a larger statistical fluctuation on the results. 
The second factor is that there appears to be a modest effect due to the population of sources that 
are detectable. For distances at which the \ac{BNS} and \ac{NSBH} systems would give the same 
percentage of observable sources there is a slight increase in the mean and standard distribution 
of \acp{SNR} of the \ac{NSBH} systems over the \ac{BNS} systems.

As noted previously the mass distribution for the black holes could be quite different from the one 
we used, with masses extending to $\gtrsim 10$\,M$_{\odot}$. We expect these would produce similar 
results to those we see in figure~\ref{fig:nsbhresults}, but again extending to further distances. 
Including the merger and ring-down phase, which we currently ignore, may also help increase 
\ac{SNR}.

@@

We have attempted to see how well we can assess the detector calibration for \ac{aLIGO} and 
\ac{AdV} using astrophysical sources. To do this we require that a \ac{sGRB}, for which it has 
been possible to measure the distance, is observed in coincidence with a \ac{CBC} signal in the 
\ac{GW} detectors. This enables us to assume a known distance to the source, and also limit the 
inclination of the source, which in turn allows us to test the consistency of the detector 
calibration. We do this by including an unknown scale factor on the true signal and noise, which we 
estimate given the data. We find that for detectable \ac{BNS} sources the uncertainty of on the 
calibration scale factor could on average be determined to $\lesssim 10\%$ of its true value for the 
\ac{aLIGO} and \ac{AdV} detectors out to 100\,Mpc. This is comparable to the hoped for level of 
calibration accuracy of the detectors. For sources at the standard single detector \ac{BNS} horizon 
distance of $\sim 450$\,Mpc the scale factor could on average be determined to within $\lesssim 
20\%$ of its true value for the \ac{aLIGO} detectors. Similar results were found for \ac{NSBH} 
sources, although sources could be observed out to higher distances.

The requirement of a coincident \ac{sGRB} with a known distance means that there will be 
considerable latency for this method of calibration assessment, but importantly this method would 
provide semi-independent consistency check that the calibration is reasonably accurate. We do note 
that this method has to make its own assumptions, including that the waveforms we use accurately 
reflect the signal, that the phase calibration does not change to much over frequency and that 
the overall calibration does not change considerably over the length of a signal. These are studies 
that could be incorporated in the future.

We imagine that a test for consistency of calibration could be performed with an ensemble of 
\ac{CBC} sources even without \ac{sGRB} counterparts, although such a method would have to 
marginalise over highly uncertain population models and may give quite weak bounds on the overall 
calibration uncertainty.

We emphasise that our work has assumed that the \ac{PSD} used in the likelihood function when 
estimating the calibration scale factor is an accurate representation of the \ac{PSD} at the time 
of the observed \ac{CBC} signal. In reality the \ac{PSD} may be estimated from a period close to, 
but not overlapping, the signal time and therefore may be slightly different 
\cite{2013PhRvD..88h4044L}. Our result would therefore be highly correlated with any 
uncertainty in the \ac{PSD} estimate, but would still offer an upper limit on the calibration 
uncertainty. However, future plans for \ac{CBC} parameter estimation include estimating the 
\ac{PSD} with the signal using BayesLine \cite{2015PhRvD..91h4034L}.

A further extension of this work would be for the calibration scale factor to be estimated as a 
function of frequency. This would require the frequency series to be broken up into sections and 
the calibration scale factor estimated for each. However, the scale factor uncertainty will 
obviously be related to the \ac{SNR} of a signal within a particular section. It could 
therefore be a job of Bayesian model selection to work out the optimal number of frequency 
sections to estimate the scale factor for. The BayesLine algorithm \cite{2015PhRvD..91h4034L}, or 
the method described in \cite{2013PhRvD..88h4044L}, may provide a natural way to perform such an
analysis. Indeed these methods, and those of \cite{Vitale:2012} are already being used for 
marginalising over calibration uncertainties, but do not attempt to estimate it.
